Multi-level generalizations of collective spin dynamics and squeezing

### 2019-10-04 collective_sun/sun_notes.tex
changed some notation and added first results for weak SOC

An driving laser addressing spin-$1/2$ fermions on a 1-D lattice will
induce the Hamiltonian
\begin{align}
  \left. H_{\t{drive}}^{(\phi)} \right|_{n=2}
  = \Omega \sum_j
  \p{e^{-i\phi j} c_{j,\up}^\dag c_{j,\dn}
    + e^{i\phi j} c_{j,\dn}^\dag c_{j,\up}},
  \label{eq:H_drive_2}
\end{align}
where $\Omega$ is a (real) driving amplitude, $j\in\mathbb{Z}$ indexes
a lattice site, and $\phi$ is the relative phase of the drive between
adjacent lattice sites.  The multi-level analogue of
\eqref{eq:H_drive_2} is
\begin{align}
  H_{\t{drive}}^{(\phi)}
  = -\Omega \sum_j \p{e^{-i\phi j} T_{1,1}^{(j)}
    + e^{i\phi j} {T_{1,1}^{(j)}}^\dag}
  = \Omega \D_{1,1}^{(\phi)},
  \label{eq:H_drive}
\end{align}
where
\begin{align}
  \D_{LM}^{(\phi)}
  \equiv \sum_j \sp{\cos\p{M\phi j} D_{LM}^{(j)}
    + \sin\p{M\phi j} D_{L,-M}^{(j)}},
  \label{eq:drive_rot}
\end{align}
is an inhomogeneously rotated collective drive operator.  We can make
this drive spatially homogenous via the gauge transformation
$c_{j\mu}^\dag \to e^{i\mu\phi j} c_{j\mu}^\dag$, which takes
\begin{align}
  H_{\t{drive}}^{(\phi)} \to H_{\t{drive}} = \Omega \D_{1,1}.
\end{align}
The SU($n$) symmetry of collective interactions implies that the
interaction Hamiltonian $H_{\t{int}}$ in \eqref{eq:H_int_drive} is
invariant under this gauge transformation, but the single-particle
Hamiltonian in \eqref{eq:H_lat} transforms as (see Appendix
\ref{sec:lat_drive})
\begin{align}
  H_{\t{lat}}
  \to H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa-\mu\phi} c_{q\mu}^\dag c_{q\mu}
  = \sum_{q,L} B_L^{(\phi,q)} D_{L,0}^{(q)},
  \label{eq:H_lat_SOC}
\end{align}
with effective inhomogeneous driving field amplitudes
\begin{align}
  B_L^{(\phi,q)} = -J f_L\p{qa} \sqrt{\f{L+1/2}{n}}
  \sum_\mu \bk{I\mu,L,0;I\mu} f_L\p{\mu\phi},
  &&
  f_L\p{\theta} \equiv
  \begin{cases}
    \cos\theta & L~\t{even} \\
    \sin\theta & L~\t{odd}
  \end{cases}.
\end{align}
Spin-orbit coupling with multi-level systems thus allows for the
simulation of synthetic gauge fields with large spin ($L>1$).  As we
will see in the following section, multi-laser drives can also
simulate synthetic large-spin fields with non-zero spin projection
($M\ne0$) onto a quantization axis.

@@

Here we derive the drive-operator ($D_{LM}$) expansion of the
spin-orbit coupled Hamiltonian $H_{\t{lat}}^{(\phi)}$ in
\eqref{eq:H_lat_SOC}.  In the basis of spin operators
$s_{\mu\mu}^{(q)} \equiv c_{q\mu}^\dag c_{q\mu}$, this Hamiltonian
takes the form
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa-\mu\phi} s_{\mu\mu}^{(q)}
  \label{eq:H_lat_spin}
\end{align}
Defining the diagonal ``spin-wave'' operators
\begin{align}
  w_{\phi+} \equiv \sum_\mu \cos\p{\mu\phi} s_{\mu\mu},
  &&
  w_{\phi-} \equiv \sum_\mu \sin\p{\mu\phi} s_{\mu\mu},
\end{align}
which are respectively even and odd under spin inversion
($\mu\to-\mu$), we can expand
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} w_{\phi+} + \sin\p{qa} w_{\phi-},
\end{align}
and resolve the identity (super-)operator $\I$ with
$\I w_{\phi,\pm} = w_{\phi,\pm}$ in the basis of the drive
operators $D_{LM}$, finding (see Appendix \ref{sec:changing_bases})
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L,M} \obk{w_{\phi+}|D_{LM}} D_{LM}
  + \sin\p{qa} \f12 \sum_{L,M} \obk{w_{\phi-}|D_{LM}} D_{LM}.
\end{align}
We now note that the drive operators $D_{LM}$ are strictly
off-diagonal for $M\ne0$, and furthermore that $D_{L,0}$ with even
(odd) $L$ is even (odd) under spin inversion, which implies
\begin{align}
  \sum_{L,M} \obk{w_{\phi+}|D_{LM}} D_{LM}
  = \sum_{L\,\t{even}} \obk{w_{\phi+}|D_{L,0}} D_{L,0},
\end{align}
and likewise with $w_{\phi-}$ and odd $L$.  In total, we can
write
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L\,\t{even}}
  \obk{w_{\phi+}|D_{L,0}} D_{L,0}
  + \sin\p{qa} \f12 \sum_{L\,\t{odd}}
  \obk{w_{\phi,-}|D_{L,0}} D_{L,0},
\end{align}
or, more compactly,
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \sum_L A_L^{(\phi,q)} D_{L,0},
  \label{eq:compact_SOC_drive}
\end{align}
with effective inhomogeneous driving field amplitudes
\begin{align}
  A_L^{(\phi,q)}
  \equiv f_L\p{qa} \f12
  \sum_\mu f_L\p{\mu\phi} \obk{s_{\mu\mu}|D_{L,0}}
  = f_L\p{qa} \sqrt{\f{L+1/2}{n}}
  \sum_\mu \bk{I\mu,L,0;I\mu} f_L\p{\mu\phi},
\end{align}
where
\begin{align}
  f_L\p{\theta} \equiv
  \begin{cases}
    \cos\theta & L~\t{even} \\
    \sin\theta & L~\t{odd}
  \end{cases}.
\end{align}
Substituting the result in \eqref{eq:compact_SOC_drive} into the
single-particle Hamiltonian in \eqref{eq:H_lat_spin} yields
\begin{align}
  H_{\t{lat}}^{(\phi)} = \sum_{q,L} B_L^{(\phi,q)} D_{L,0}^{(q)},
  &&
  B_L^{(\phi,q)} \equiv -J A_L^{(\phi,q)}.
\end{align}

@@

Here we provide the coefficients $\Omega_{LM}^{(\phi,j)}$ of the
effective Raman driving Hamiltonian
\begin{align}
  H_{\t{Raman}}^{(\phi)}
  = \sum_j \sum_{\substack{L\in\set{0,1,2}\\\abs{M}\le L}}
  \Omega_{LM}^{(\phi,j)} D_{LM}^{(j)}
\end{align}
considered in Section \ref{sec:drive_raman}.  These coefficients are
computed by evaluating products of transition operators in
\eqref{eq:raman_0}--\eqref{eq:raman_2}, using both the structure
constants of the transition operator algebra (derived in Appendix
\ref{sec:trans_prod}) and the fact that
$T_{LM}^\dag = \p{-1}^M T_{L,-M}$.  For clarity, if any coefficient
$\Omega_{LM}^{(\phi,j)}$ is independent of $\phi$, we drop the
explicit dependence on $\phi$, i.e.~taking
$\Omega_{LM}^{(\phi,j)}\to\Omega_{LM}$.
\begin{align}
  \Omega_{0,0} \equiv -\f1{\sqrt{2n}} \sum_m\f{\Omega_m^2}{\Delta},
  &&
  \Omega_{1,0} \equiv \sqrt{\f{3}{2n\p{n^2-1}}}~
  \f{\Omega_+^2 - \Omega_-}{\Delta},
  \label{eq:O_01_0}
\end{align}
\begin{align}
  \Omega_{2,0} \equiv \sqrt{\f{n^2-4}{10n\p{n^2-1}}}~
  \f{\Omega_+ + \Omega_- - 2\Omega_0}{\Delta},
  \label{eq:O_2_0}
\end{align}
\begin{align}
  \Omega_{L,1}^{(\phi,j)}
  \equiv \sum_{s\in\set{\pm1}} \omega_{Ls} \cos\p{\phi_s j},
  &&
  \Omega_{L,-1}^{(\phi,j)}
  \equiv \sum_{s\in\set{\pm1}} s \omega_{Ls} \sin\p{\phi_s j},
\end{align}
\begin{align}
  \Omega_{2,2}^{(\phi,j)}
  \equiv \omega_{2,2} \cos\p{\sp{\phi_+-\phi_-}j},
  &&
  \Omega_{2,-2}^{(\phi,j)}
  \equiv \omega_{2,2} \sin\p{\sp{\phi_+-\phi_-}j},
\end{align}
\begin{align}
  \omega_{1\pm} \equiv \f{\sqrt3}{\sqrt{n\p{n^2-1}}}~
  \f{\Omega_0 \Omega_\pm}{\Delta},
  &&
  \omega_{2\pm} \equiv \pm \f{\sqrt{3/5}}{n\p{n^2-1}}
  \sqrt{\f{\p{n+2}!}{\p{n-3}!}}~ \f{\Omega_0\Omega_\pm}{\Delta},
\end{align}
\begin{align}
  \omega_{2,2} \equiv -\f{\sqrt{6/5}}{n\p{n^2-1}}
  \sqrt{\f{\p{n+2}!}{\p{n-3}!}}~ \f{\Omega_+\Omega_-}{\Delta}.
\end{align}
Here $\Omega_m$, $\phi_m$, and $\Delta$ are respectively the (real)
driving amplitudes, phases, and detuning that appear in the ``bare''
Raman drive in \eqref{eq:raman_bare}, and by choice of gauge we set
$\phi_-=0$, or equivalently re-define $\phi_m-\phi_0\to\phi_m$.  If
the driving lasers are phase-matched with $\phi_s=s\phi$, such that a
particle on site $j$ acquires a phase $\mp\phi j$ for every nuclear
spin transition $\mu\to\mu\pm1$, then the drive
$H_{\t{Raman}}^{(\phi)}$ takes the form
\begin{align}
  \left. H_{\t{Raman}}^{(\phi)} \right|_{\phi_s=s\phi}
  = \sum_{\substack{L\le2\\0\le M\le L}} \Omega_{LM} \D_{LM}^{(\phi)},
\end{align}
with inhomogeneously rotated collective drive operators
\begin{align}
  \D_{LM}^{(\phi)}
  \equiv \sum_j \sp{\cos\p{M\phi j} D_{LM}^{(j)}
    + \sin\p{M\phi j} D_{L,-M}^{(j)}},
\end{align}
and uniform coefficients given by \eqref{eq:O_01_0}, \eqref{eq:O_2_0},
and
\begin{align}
  \Omega_{1,-1} = \Omega_{2,-1} = \Omega_{2,-2} \equiv 0,
  &&
  \Omega_{L,1} \equiv \sum_{s\in\set{\pm1}} \omega_{Ls},
  &&
  \Omega_{2,2} \equiv \omega_{2,2}.
\end{align}

>>>>

An driving laser addressing spin-$1/2$ fermions on a 1-D lattice will
induce the Hamiltonian
\begin{align}
  \left. H_{\t{drive}}^{(\phi)} \right|_{n=2}
  = \Omega \sum_j
  \p{e^{-i\phi j} c_{j,\up}^\dag c_{j,\dn}
    + e^{i\phi j} c_{j,\dn}^\dag c_{j,\up}},
  \label{eq:H_drive_2}
\end{align}
where $\Omega$ is a (real) driving amplitude, $j\in\mathbb{Z}$ indexes
a lattice site, and $\phi$ is the relative phase of the drive between
adjacent lattice sites.  The multi-level analogue of
\eqref{eq:H_drive_2} is
\begin{align}
  H_{\t{drive}}^{(\phi)}
  = -\Omega \sum_j \p{e^{-i\phi j} T_{1,1}^{(j)}
    + e^{i\phi j} {T_{1,1}^{(j)}}^\dag}
  = \Omega \D_{1,1}^{(\phi)},
  \label{eq:H_drive}
\end{align}
where
\begin{align}
  \D_{LM}^{(\phi)}
  \equiv \sum_j \sp{\cos\p{M\phi j} D_{LM}^{(j)}
    + \sin\p{M\phi j} D_{L,-M}^{(j)}},
  \label{eq:drive_rot}
\end{align}
is an inhomogeneously rotated collective drive operator.  We can make
this drive spatially homogenous via the gauge transformation
$c_{j\mu}^\dag \to e^{i\mu\phi j} c_{j\mu}^\dag$, which takes
\begin{align}
  H_{\t{drive}}^{(\phi)} \to H_{\t{drive}} = \Omega \D_{1,1}.
\end{align}
The SU($n$) symmetry of collective interactions implies that the
interaction Hamiltonian $H_{\t{int}}$ in \eqref{eq:H_int_drive} is
invariant under this gauge transformation, but the single-particle
Hamiltonian in \eqref{eq:H_lat} transforms as (see Appendix
\ref{sec:lat_drive})
\begin{align}
  H_{\t{lat}}
  \to H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa-\mu\phi} c_{q\mu}^\dag c_{q\mu}
  = \sum_{q,L} B_L^{(\phi,q)} D_{L,0}^{(q)},
  \label{eq:H_lat_SOC}
\end{align}
with effective inhomogeneous driving field amplitudes
\begin{align}
  B_L^{(\phi,q)} = -J f_L\p{qa} W_L^{(\phi)},
\end{align}
where
\begin{align}
  W_L^{(\phi)} \equiv \sqrt{\f{L+1/2}{n}}
  \sum_\mu \bk{I\mu,L,0;I\mu} f_L\p{\mu\phi}
  &&
  f_L\p{\theta} \equiv
  \begin{cases}
    \cos\theta & L~\t{even} \\
    \sin\theta & L~\t{odd}
  \end{cases}.
\end{align}
Spin-orbit coupling with multi-level systems thus allows for the
simulation of synthetic gauge fields with large spin ($L>1$).  As we
will see in the following section, multi-laser drives can also
simulate synthetic large-spin fields with non-zero spin projection
($M\ne0$) onto a quantization axis.

In the regime of weak SOC with $\abs{\phi}\ll1$, we find that
\begin{align}
  W_L^{(\phi)} = c_L \sqrt{n_L\p{L+1/2}} \phi^L + O\p{\phi^{L+2}},
  &&
  n_L \equiv \prod_{\ell=-L}^L\p{n-\ell},
\end{align}
for
\begin{align}
  c_0 = 1,
  &&
  c_1 = \f16,
  &&
  c_2 = \f1{30},
  &&
  c_3 = \f1{140},
  &&
  c_4 = \f1{630},
  &&
  c_5 = \f1{2772}.
\end{align}

@@

Here we derive the drive-operator ($D_{LM}$) expansion of the
spin-orbit coupled Hamiltonian $H_{\t{lat}}^{(\phi)}$ in
\eqref{eq:H_lat_SOC}.  In the basis of spin operators
$s_{\mu\mu}^{(q)} \equiv c_{q\mu}^\dag c_{q\mu}$, this Hamiltonian
takes the form
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa-\mu\phi} s_{\mu\mu}^{(q)}
  \label{eq:H_lat_spin}
\end{align}
Defining the diagonal ``spin-wave'' operators
\begin{align}
  w_{\phi+} \equiv \sum_\mu \cos\p{\mu\phi} s_{\mu\mu},
  &&
  w_{\phi-} \equiv \sum_\mu \sin\p{\mu\phi} s_{\mu\mu},
\end{align}
which are respectively even and odd under spin inversion
($\mu\to-\mu$), we can expand
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} w_{\phi+} + \sin\p{qa} w_{\phi-},
\end{align}
and resolve the identity (super-)operator $\I$ with
$\I w_{\phi\pm} = w_{\phi\pm}$ in the basis of the drive operators
$D_{LM}$ (see Appendix \ref{sec:changing_bases}), finding
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L,M} \obk{w_{\phi+}|D_{LM}} D_{LM}
  + \sin\p{qa} \f12 \sum_{L,M} \obk{w_{\phi-}|D_{LM}} D_{LM}.
\end{align}
We now note that the drive operators $D_{LM}$ are strictly
off-diagonal for $M\ne0$, and furthermore that $D_{L,0}$ with even
(odd) $L$ is even (odd) under spin inversion, which implies
\begin{align}
  \sum_{L,M} \obk{w_{\phi+}|D_{LM}} D_{LM}
  = \sum_{L\,\t{even}} \obk{w_{\phi+}|D_{L,0}} D_{L,0},
\end{align}
and likewise with $w_{\phi-}$ and odd $L$.  In total, we can
write
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L\,\t{even}}
  \obk{w_{\phi+}|D_{L,0}} D_{L,0}
  + \sin\p{qa} \f12 \sum_{L\,\t{odd}}
  \obk{w_{\phi-}|D_{L,0}} D_{L,0},
\end{align}
or, more compactly,
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \sum_L f_L\p{qa} W_L^{(\phi)} D_{L,0},
  &&
  f_L\p{\theta} \equiv
  \begin{cases}
    \cos\theta & L~\t{even} \\
    \sin\theta & L~\t{odd}
  \end{cases},
  \label{eq:compact_SOC_drive}
\end{align}
with effective driving field amplitudes
\begin{align}
  W_L^{(\phi)}
  \equiv \f12 \obk{w_{\phi,\p{-1}^L}|D_{L,0}}
  = \f12 \sum_\mu f_L\p{\mu\phi} \obk{s_{\mu\mu}|D_{L,0}}
  = \sqrt{\f{L+1/2}{n}}
  \sum_\mu \bk{I\mu,L,0;I\mu} f_L\p{\mu\phi}.
\end{align}
Substituting the result in \eqref{eq:compact_SOC_drive} into the
single-particle Hamiltonian in \eqref{eq:H_lat_spin} yields
\begin{align}
  H_{\t{lat}}^{(\phi)} = \sum_{q,L} B_L^{(\phi,q)} D_{L,0}^{(q)},
  &&
  B_L^{(\phi,q)} \equiv -J f_L\p{qa} W_L^{(\phi)}.
\end{align}

@@

Here we provide the coefficients $\Omega_{LM}^{(\phi,j)}$ of the
effective Raman driving Hamiltonian
\begin{align}
  H_{\t{Raman}}^{(\phi)}
  = \sum_j \sum_{\substack{L\in\set{0,1,2}\\\abs{M}\le L}}
  \Omega_{LM}^{(\phi,j)} D_{LM}^{(j)}
\end{align}
considered in Section \ref{sec:drive_raman}.  These coefficients are
computed by evaluating products of transition operators in
\eqref{eq:raman_0}--\eqref{eq:raman_2}, using both the structure
constants of the transition operator algebra (derived in Appendix
\ref{sec:trans_prod}) and the fact that
$T_{LM}^\dag = \p{-1}^M T_{L,-M}$.  For clarity, if any coefficient
$\Omega_{LM}^{(\phi,j)}$ is independent of $\phi$, we drop the
explicit dependence on $\phi$, i.e.~taking
$\Omega_{LM}^{(\phi,j)}\to\Omega_{LM}$.
\begin{align}
  \Omega_{0,0} \equiv -\f1{\sqrt{2}}~ \f1{\sqrt{n}}~
  \sum_m\f{\Omega_m^2}{\Delta},
  &&
  \Omega_{1,0} \equiv \sqrt{\f32}~ \f1{\sqrt{n_1}}~
  \f{\Omega_+^2 - \Omega_-^2}{\Delta},
  \label{eq:O_01_0}
\end{align}
\begin{align}
  \Omega_{2,0} \equiv \f1{\sqrt{10}}~ \f{\sqrt{n_2}}{n_1}~
  \f{\Omega_+ + \Omega_- - 2\Omega_0}{\Delta},
  \label{eq:O_2_0}
\end{align}
\begin{align}
  \Omega_{L,1}^{(\phi,j)}
  &\equiv \sum_{s\in\set{\pm1}} \omega_{Ls} \cos\p{\phi_s j},
  &
  \Omega_{L,-1}^{(\phi,j)}
  &\equiv \sum_{s\in\set{\pm1}} s \omega_{Ls} \sin\p{\phi_s j}, \\
  \Omega_{2,2}^{(\phi,j)}
  &\equiv \omega_{2,2} \cos\p{\sp{\phi_+-\phi_-}j},
  &
  \Omega_{2,-2}^{(\phi,j)}
  &\equiv \omega_{2,2} \sin\p{\sp{\phi_+-\phi_-}j}, \\
  \omega_{1\pm}
  &\equiv \sqrt{3} ~\f1{\sqrt{n_1}}~
  \f{\Omega_0 \Omega_\pm}{\Delta},
  &
  \omega_{2\pm}
  &\equiv \pm \sqrt{\f35}~ \f{\sqrt{n_2}}{n_1}~
  \f{\Omega_0\Omega_\pm}{\Delta}, \\
  \omega_{2,2}
  &\equiv -\sqrt{\f65}~ \f{\sqrt{n_2}}{n_1}~
  \f{\Omega_+\Omega_-}{\Delta},
  &
  n_L
  &\equiv \prod_{\ell=-L}^L \p{n-\ell}.
\end{align}
Here $\Omega_m$, $\phi_m$, and $\Delta$ are respectively the (real)
driving amplitudes, phases, and detuning that appear in the ``bare''
Raman drive in \eqref{eq:raman_bare}, and by choice of gauge we set
$\phi_-=0$, or equivalently re-define $\phi_m-\phi_0\to\phi_m$.  If
the driving lasers are phase-matched with $\phi_s=s\phi$, such that a
particle on site $j$ acquires a phase $\mp\phi j$ for every nuclear
spin transition $\mu\to\mu\pm1$, then the drive
$H_{\t{Raman}}^{(\phi)}$ takes the form
\begin{align}
  \left. H_{\t{Raman}}^{(\phi)} \right|_{\phi_s=s\phi}
  = \sum_{\substack{L\le2\\0\le M\le L}} \Omega_{LM} \D_{LM}^{(\phi)},
\end{align}
with inhomogeneously rotated collective drive operators
\begin{align}
  \D_{LM}^{(\phi)}
  \equiv \sum_j \sp{\cos\p{M\phi j} D_{LM}^{(j)}
    + \sin\p{M\phi j} D_{L,-M}^{(j)}},
\end{align}
and uniform coefficients given by \eqref{eq:O_01_0}, \eqref{eq:O_2_0},
and
\begin{align}
  \Omega_{1,-1} = \Omega_{2,-1} = \Omega_{2,-2} \equiv 0,
  &&
  \Omega_{L,1} \equiv \sum_{s\in\set{\pm1}} \omega_{Ls},
  &&
  \Omega_{2,2} \equiv \omega_{2,2}.
\end{align}
